---
execute:
  warning: false
  message: false
---

# Workflow: code style {#sec-code-style}

If you haven't installed the styler package, run this once in the Console:

```{r}
#| eval: false
install.packages("styler")
```

```{r}
library(tidyverse)
library(nycflights13)
```

The code in the "Strive for" and "Avoid" examples below is only shown, not run.

## Names

Variable names (those created by `<-` and those created by `mutate()`) should use only lowercase letters, numbers, and `_`. Use `_` to separate words within a name.

```{r}
#| eval: false
# Strive for:
short_flights <- flights |>
  filter(air_time < 60)

# Avoid:
SHORTFLIGHTS <- flights |>
  filter(air_time < 60)
```

As a general rule of thumb, it's better to prefer long, descriptive names that are easy to understand rather than concise names that are fast to type.

## Spaces

Put spaces on either side of mathematical operators apart from `^` (i.e. `+`, `-`, `==`, `<`, ...), and around the assignment operator (`<-`).

```{r}
#| eval: false
# Strive for
z <- (a + b)^2 / d

# Avoid
z<-( a + b ) ^ 2/d
```

Don't put spaces inside or outside parentheses for regular function calls. Always put a space after a comma, just like in standard English.

```{r}
#| eval: false
# Strive for
mean(x, na.rm = TRUE)

# Avoid
mean (x ,na.rm=TRUE)
```

You can add extra spaces if it improves alignment. For example, if you're creating multiple variables in `mutate()`, you might add spaces so that all the `=` line up. This makes the code easier to skim.

```{r}
#| eval: false
flights |>
  mutate(
    speed      = distance / air_time,
    dep_hour   = dep_time %/% 100,
    dep_minute = dep_time %%  100
  )
```

## Pipes

`|>` should always have a space before it and should typically be the last thing on a line.

```{r}
#| eval: false
# Strive for
flights |>
  filter(!is.na(arr_delay), !is.na(tailnum)) |>
  count(dest)

# Avoid
flights|>filter(!is.na(arr_delay), !is.na(tailnum))|>count(dest)
```

If the function you're piping into has named arguments (like `mutate()` or `summarize()`), put each argument on a new line. If the function doesn't have named arguments (like `select()` or `filter()`), keep everything on one line unless it doesn't fit, in which case put each argument on its own line.

```{r}
#| eval: false
# Strive for
flights |>
  group_by(tailnum) |>
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

# Avoid
flights |>
  group_by(
    tailnum
  ) |>
  summarize(delay = mean(arr_delay, na.rm = TRUE), n = n())
```

After the first step of the pipeline, indent each line by two spaces. RStudio will automatically put the spaces in for you after a line break following a `|>`. If you're putting each argument on its own line, indent by an extra two spaces. Make sure `)` is on its own line, and un-indented to match the horizontal position of the function name.

```{r}
#| eval: false
# Strive for
flights |>
  group_by(tailnum) |>
  summarize(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )

# Avoid: arguments and closing parenthesis indented too far
flights |>
  group_by(tailnum) |>
  summarize(
      delay = mean(arr_delay, na.rm = TRUE),
      n = n()
    )
```

## ggplot2

The same basic rules that apply to the pipe also apply to ggplot2; just treat `+` the same way as `|>`.

```{r}
flights |>
  group_by(month) |>
  summarize(delay = mean(arr_delay, na.rm = TRUE)) |>
  ggplot(aes(x = month, y = delay)) +
  geom_point() +
  geom_line()
```

```{r}
flights |>
  group_by(dest) |>
  summarize(
    distance = mean(distance),
    speed = mean(distance / air_time, na.rm = TRUE)
  ) |>
  ggplot(aes(x = distance, y = speed)) +
  geom_smooth(
    method = "loess",
    span = 0.5,
    se = FALSE,
    color = "red",
    linewidth = 4
  ) +
  geom_point()
```

## Sectioning comments

*(Notes to come.)*

## Exercises

**1. Restyle the following pipelines following the guidelines above.**

The messy versions:

```{r}
#| eval: false
flights|>filter(dest=="IAH")|>group_by(year,month,day)|>summarize(n=n(),
                                                                  delay=mean(arr_delay,na.rm=TRUE))|>filter(n>10)

flights|>filter(carrier=="UA",dest%in%c("IAH","HOU"),sched_dep_time>
                  0900,sched_arr_time<2000)|>group_by(flight)|>summarize(delay=mean(
                    arr_delay,na.rm=TRUE),cancelled=sum(is.na(arr_delay)),n=n())|>filter(n>10)
```

The clean versions:

```{r}
flights |>
  filter(dest == "IAH") |>
  group_by(year, month, day) |>
  summarise(n = n(), delay = mean(arr_delay, na.rm = TRUE)) |>
  filter(n > 10)
```

```{r}
flights |>
  filter(
    carrier == "UA",
    dest %in% c("IAH", "HOU"),
    sched_dep_time > 0900,
    sched_arr_time < 2000
  ) |>
  group_by(flight) |>
  summarise(
    delay = mean(arr_delay, na.rm = TRUE),
    cancelled = sum(is.na(arr_delay)),
    n = n()
  ) |>
  filter(n > 10)
```
