---
execute:
  warning: false
  message: false
---

# Data transformation {#sec-data-transformation}

## Introduction

### Prerequisites

We'll be using **dplyr** with **nycflights13**, a package with multiple datasets: `flights`, `weather`, `airlines`, `airports` and `planes`.

If you haven't installed nycflights13 yet, run this once in the Console:

```{r}
#| eval: false
install.packages("nycflights13")
```

```{r}
library(tidyverse)
library(nycflights13)
library(palmerpenguins)
```

### nycflights13

```{r}
nrow(flights)
head(flights)
head(weather)
head(planes)
```

```{r}
#| eval: false
View(flights)
?flights
```

```{r}
flights
glimpse(flights)
```

Setting `width = Inf` basically tells R "I don't care how wide it is, show me everything":

```{r}
print(flights, width = Inf)
```

- `width` = how wide the output is allowed to be
- `Inf` = infinity = no limit

### dplyr basics

dplyr uses verbs (functions). They differ, but have these things in common:

1.  The first argument is always a data frame.
2.  The subsequent arguments typically describe which columns to operate on, using the variable names (without quotes).
3.  The output is always a new data frame.

Complex problems require multiple verbs, so we use the pipe (`|>`). In brief, the pipe takes the thing on its left and passes it along to the function on its right.

An example (I am not familiar with this yet, lol):

```{r}
flights |>
  filter(dest == "IAH") |>
  group_by(year, month, day) |>
  summarize(
    arr_delay = mean(arr_delay, na.rm = TRUE)
  )
```

The same idea for Miami:

```{r}
flights |>
  filter(dest == "MIA") |>
  group_by(year, month, day) |>
  summarize(
    arr_delay = mean(arr_delay, na.rm = TRUE)
  )
```

dplyr's verbs are organized into four groups based on what they operate on:

- rows
- columns
- tables
- groups

## Rows

### filter()

`filter()` keeps rows based on the values of the columns. It lets you say "I only want to see rows where a certain condition is true" and throws away all the rest.

The first argument is always the data frame. The second and later arguments are the conditions that must be true. For instance, flights that departed more than 120 minutes late:

```{r}
flights |>
  filter(dep_delay > 120)
```

As well as `>` (greater than), you can use `>=` (greater than or equal to), `<` (less than), `<=` (less than or equal to), `==` (equal to), and `!=` (not equal to).

You can also combine conditions with `&` or `,` to mean "and" (check for both conditions), or with `|` to mean "or" (check for either condition).

Flights that departed on January 1:

```{r}
flights |>
  filter(month == 1, day == 1)
```

or

```{r}
flights |>
  filter(month == 1 & day == 1)
```

Flights that departed in January or February (`|` means "or"):

```{r}
flights |>
  filter(month == 1 | month == 2)
```

When you're combining `|` and `==`, there is a shortcut: `%in%`. It keeps rows where the variable equals one of the values on the right. Instead of the code above, you can write:

```{r}
flights |>
  filter(month %in% c(1, 2))
```

```{r}
flights |>
  filter(month %in% c(1, 2, 3, 4))
```

Flights going to Houston OR Chicago:

```{r}
flights |>
  filter(dest %in% c("IAH", "ORD"))
```

Flights of United Airlines and American Airlines:

```{r}
flights |>
  filter(carrier %in% c("UA", "AA"))
```

It reads as "keep rows where `month` is IN this list of values". Also note that numbers never get quotes, but text always does.

To save the new data frame that dplyr has created, you need to assign it with `<-`:

```{r}
jan1 <- flights |>
  filter(month == 1, day == 1)

jan1
```

It works on other data too:

```{r}
penguins |>
  filter(species == "Adelie")
```

Saving a data frame to a file:

```{r}
#| eval: false
write_csv(jan1, "jan2.csv")
```

How to commit to GitHub in the Terminal:

``` bash
git add .
git commit -m "describe what you did"
git push
```

### Common mistakes

*(Notes to come.)*

### arrange()

`arrange()` sorts your data from lowest to highest (or highest to lowest) based on a column you choose.

If you provide more than one column name, each additional column is used to break ties in the values of the preceding columns. For instance:

```{r}
flights |>
  arrange(year, month, day, dep_time)
```

Each column only kicks in when the previous one results in a tie.

```{r}
dest_filter <- flights |>
  arrange(year, month, dest)
```

This sorts flights chronologically and then alphabetically by destination within each month. You'd see all January flights going to destinations starting with A, B, C and so on.

```{r}
#| eval: false
write.csv(dest_filter, "dest_filter.csv")
```

Which flights left earliest in the day?

```{r}
flights |>
  arrange(dep_time)
```

To re-order by a column in descending (big-to-small) order, use `desc()`. This orders from most to least delayed:

```{r}
flights |>
  arrange(desc(dep_delay))
```

Sort flights chronologically by year, month and day, and within each day show the most delayed flights first:

```{r}
flights |>
  arrange(year, month, day, desc(dep_delay))
```

### distinct()

`distinct()` removes duplicate rows from the data, keeping only unique rows.

```{r}
flights |>
  distinct()
```

The difference from `filter()`:

```{r}
# keep only Houston flights
flights |>
  filter(dest == "IAH")
```

```{r}
# keep only unique destinations
flights |>
  distinct(dest)
```

```{r}
flights |>
  distinct(dest) |>
  nrow()
```

```{r}
penguins |>
  distinct(bill_depth_mm) |>
  nrow()
```

Finding unique pairs of `dest` and `origin`:

```{r}
flights |>
  distinct(dest, origin)
```

```{r}
flights |>
  distinct(dest, origin, carrier)
```

`distinct()` removes duplicate **combinations**, not necessarily duplicate values in each column individually. So `distinct(origin, dest)` doesn't give unique origins and unique destinations separately; it gives unique *pairs* of them.

To keep the other columns when filtering for unique rows, use the `.keep_all = TRUE` option:

```{r}
flights |>
  distinct(dest, origin, carrier, .keep_all = TRUE)
```

```{r}
flights |>
  distinct(origin, dest, .keep_all = TRUE) |>
  nrow()
```

`.keep_all = TRUE` is just saying "yes, I want unique origin-dest combinations, but don't throw away the rest of my data".

`distinct()` only shows you the unique combinations. It doesn't tell you how many times each one appears:

```{r}
flights |>
  distinct(origin, dest)
```

`count()` tells you how many times each combination appears:

```{r}
flights |>
  count(origin, dest)
```

Adding `sort = TRUE` arranges by most common first:

```{r}
flights |>
  count(origin, dest, sort = TRUE)
```

```{r}
flights |>
  count(dest, sort = TRUE)
```

### Exercises

**1. In a single pipeline for each condition, find all flights that meet the condition.**

*Had an arrival delay of two or more hours:*

```{r}
flights |>
  filter(arr_delay >= 120) |>
  arrange(desc(arr_delay))
```

*Flew to Houston (IAH or HOU):*

```{r}
flights |>
  filter(dest %in% c("IAH", "HOU"))
```

or

```{r}
flights |>
  filter(dest == "IAH" | dest == "HOU")
```

*Were operated by United, American, or Delta:*

```{r}
flights |>
  filter(carrier %in% c("UA", "AA", "DL"))
```

*Departed in summer (July, August, and September):*

```{r}
flights |>
  filter(month %in% c(7, 8, 9))
```

*Arrived more than two hours late but didn't leave late:*

```{r}
flights |>
  filter(arr_delay > 120, dep_delay == 0)
```

Even better:

```{r}
flights |>
  filter(arr_delay >= 120 & dep_delay <= 0)
```

Using `dep_delay <= 0` is better because "didn't leave late" also includes flights that left early (a negative delay).

*Were delayed by at least an hour, but made up over 30 minutes in flight:*

```{r}
flights |>
  filter(dep_delay >= 60, dep_delay - arr_delay > 30)
```

Making up time means the arrival delay was less than the departure delay by more than 30 minutes.

**2. Sort `flights` to find the flights with the longest departure delays. Find the flights that left earliest in the morning.**

```{r}
flights |>
  arrange(desc(dep_delay))
```

```{r}
flights |>
  arrange(dep_time)
```

or

```{r}
flights |>
  arrange(time_hour)
```

`time_hour` might be a better column in some ways, because it stores the time properly as a real date-time object, without the gaps that clock-style times like `dep_time` have.

**3. Sort `flights` to find the fastest flights. (Hint: try including a math calculation inside of your function.)**

```{r}
flights |>
  arrange(desc(distance / air_time))
```

Or, if "fastest" means shortest flight duration:

```{r}
flights |>
  arrange(air_time)
```

**4. Was there a flight on every day of 2013?**

```{r}
flights |>
  distinct(year, month, day) |>
  count()
```

or

```{r}
flights |>
  distinct(year, month, day) |>
  nrow()
```

**5. Which flights traveled the farthest distance? Which traveled the least distance?**

```{r}
flights |>
  arrange(desc(distance))
```

and

```{r}
flights |>
  arrange(distance)
```

**6. Does it matter what order you used `filter()` and `arrange()` if you're using both? Why/why not? Think about the results and how much work the functions would have to do.**

The order doesn't matter for the final result, because `filter()` changes which rows are included, while `arrange()` only changes their order. However, the order can matter for efficiency. It is usually better to `filter()` first, because then `arrange()` has fewer rows to sort.

## Columns

### mutate()

`mutate()` lets you create new columns using math or logic on columns that already exist.

To compute the `gain` (how much time a delayed flight made up in the air) and the `speed` in miles per hour:

```{r}
flights |>
  mutate(gain = dep_delay - arr_delay,
         speed = distance / air_time * 60)
```

Create a TRUE/FALSE column: was the flight late?

```{r}
flights |>
  mutate(is_late = dep_delay > 0)
```

Minutes to hours:

```{r}
flights |>
  mutate(air_time_hours = air_time / 60)
```

```{r}
flights |>
  mutate(is_late = dep_delay > 0) |>
  filter(is_late)
```

By default, `mutate()` adds new columns on the right-hand side of your dataset, which makes it difficult to see what's happening above. Use the `.before` argument to add the variables to the left-hand side instead:

```{r}
flights |>
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60,
    .before = 1
  )
```

You can also use `.after` to add after a variable. In both `.before` and `.after` you can use the variable name instead of a position.

```{r}
flights |>
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60,
    .after = month
  )
```

You can control which variables are kept with the `.keep` argument. A useful value is `"used"`, which keeps only the columns that were involved in or created by the `mutate()` step:

```{r}
flights |>
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .keep = "used"
  )
```

### select()

`select()` lets you rapidly zoom in on a useful subset, using operations based on the names of the variables.

Select columns by name:

```{r}
flights |>
  select(year, day, month)
```

Select all columns between `year` and `carrier` (inclusive):

```{r}
flights |>
  select(year:carrier)
```

Select all columns except those from `year` to `dep_delay` (inclusive):

```{r}
flights |>
  select(!year:dep_delay)
```

Select all columns that are characters:

```{r}
flights |>
  select(where(is.character))
```

You can rename variables as you `select()` them by using `=`. The new name appears on the left-hand side of the `=`, and the old variable appears on the right-hand side:

```{r}
flights |>
  select(tail_num = tailnum)
```

```{r}
flights |>
  select(tailnum)
```

### rename()

It always takes the form `rename(new_name = old_name)`.

If you want to keep all the existing variables and just rename a few, use `rename()` instead of `select()`:

```{r}
flights |>
  rename(tail_number = tailnum,
         destination = dest)
```

### relocate()

`relocate()` moves variables around. You might want to collect related variables together or move important variables to the front. By default, it moves variables to the front:

```{r}
flights |>
  relocate(time_hour, air_time)
```

You can also specify where to put them using the `.before` and `.after` arguments, just like in `mutate()`:

```{r}
flights |>
  relocate(year:dep_time, .after = time_hour)
```

```{r}
flights |>
  relocate(starts_with("arr"), .before = year)
```

```{r}
flights |>
  relocate(starts_with("dep"), .after = year)
```

```{r}
flights |>
  relocate(ends_with("time"), .before = year)
```

### Exercises

**1. Compare `dep_time`, `sched_dep_time`, and `dep_delay`. How would you expect those three numbers to be related?**

```{r}
flights |>
  select(dep_time, sched_dep_time, dep_delay)
```

`dep_delay` is the difference between the actual departure time and the scheduled departure time. Knowing any two of the three variables should conceptually let you work out the third, although `dep_time` and `sched_dep_time` are stored as clock times (like 517 for 5:17) rather than as simple numbers of minutes.

**2. Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from `flights`.**

```{r}
flights |>
  select(dep_time, dep_delay, arr_time, arr_delay)
```

```{r}
flights |>
  select(starts_with("dep"), starts_with("arr"))
```

```{r}
flights |>
  select(contains("dep_"), contains("arr_"))
```

**3. What happens if you specify the name of the same variable multiple times in a `select()` call?**

```{r}
flights |>
  select(year, day, arr_time, year)
```

`select()` doesn't duplicate the variable, and it respects the order in which you first select the columns.

**4. What does the `any_of()` function do? Why might it be helpful in conjunction with this vector?**

```{r}
variables <- c("year", "month", "day", "dep_delay", "arr_delay")
```

`any_of()` lets you use a character vector of column names inside `select()` to select those columns. It's useful when the names are stored in a vector rather than typed directly into `select()`, and it can safely ignore names that don't exist.

For example, with a vector that includes a column that doesn't exist:

```{r}
variables <- c(
  "year",
  "month",
  "day",
  "dep_delay",
  "arr_delay",
  "banana"
)

flights |>
  select(any_of(variables))
```

Without `any_of()`, the missing column causes an error:

```{r}
#| error: true
flights |>
  select(variables)
```

**5. Does the result of running the following code surprise you? How do the select helpers deal with upper and lower case by default? How can you change that default?**

```{r}
flights |>
  select(contains("TIME"))
```

```{r}
flights |>
  select(starts_with("DEP"))
```

The `select()` helper functions such as `contains()` and `starts_with()` are **case-insensitive** by default.

To change that default, use `ignore.case = FALSE`:

```{r}
flights |>
  select(contains("TIME", ignore.case = FALSE))
```

Now R looks for "TIME" exactly as written. But the actual `flights` columns contain "time" in lowercase, so nothing matches:

```{r}
flights |>
  select(contains("time", ignore.case = FALSE))
```

`ignore.case = FALSE` tells R to be case sensitive.

**6. Rename `air_time` to `air_time_min` to indicate units of measurement and move it to the beginning of the data frame.**

```{r}
flights |>
  rename(air_time_min = air_time) |>
  relocate(air_time_min)
```

**7. Why doesn't the following work, and what does the error mean?**

```{r}
#| error: true
flights |>
  select(tailnum) |>
  arrange(arr_delay)
```

`select()` creates a new data frame that only has `tailnum`, so `arr_delay` no longer exists and there is nothing to arrange by.

## The pipe

*(Notes to come.)*

## Groups

Grouping divides a dataset into groups for meaningful analysis.

### group_by()

```{r}
flights |>
  group_by(month)
```

`group_by()` doesn't change the data, but if you look closely at the output, it says the data is "grouped by" month (`Groups: month [12]`). This means subsequent operations will now work "by month".

### summarize()

If it is used to calculate a single summary statistic, `summarize()` reduces the data frame to a single row for each group.

```{r}
flights |>
  group_by(month) |>
  summarise(avg_delay = mean(dep_delay))
```

This throws up `NA`s because there are some missing values in the dataset. Use `na.rm = TRUE` to remove them:

```{r}
flights |>
  group_by(month) |>
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE))
```

```{r}
flights |>
  group_by(origin) |>
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE))
```

There can be multiple summaries in a single `summarise()` call, for instance `n()`, which returns the number of rows in each group:

```{r}
flights |>
  group_by(month) |>
  summarise(
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )
```

```{r}
flights |>
  group_by(month) |>
  summarise(
    avg_distance = mean(distance, na.rm = TRUE),
    n = n()
  )
```

### The slice\_ functions

These mainly work with rows based on position. For example, the first five rows of `flights`:

```{r}
flights |>
  slice(1:5)
```

or row 32:

```{r}
flights |>
  slice(32)
```

They also let you extract specific rows within each group:

- `df |> slice_head(n = 1)` takes the first row from each group.
- `df |> slice_tail(n = 1)` takes the last row in each group.
- `df |> slice_min(x, n = 1)` takes the row with the smallest value of column `x`.
- `df |> slice_max(x, n = 1)` takes the row with the largest value of column `x`.
- `df |> slice_sample(n = 1)` takes one random row.

To find the flights that are most delayed upon arrival at each destination:

```{r}
flights |>
  group_by(dest) |>
  slice_max(arr_delay, n = 1) |>
  relocate(dest)
```

Note that there are 105 destinations but we get 108 rows here. What's up? `slice_min()` and `slice_max()` keep tied values, so `n = 1` means "give us all rows with the highest value". If you want exactly one row per group, set `with_ties = FALSE`:

```{r}
flights |>
  group_by(dest) |>
  slice_max(arr_delay, n = 1, with_ties = FALSE)
```

To find the flights most delayed at departure at each origin:

```{r}
flights |>
  group_by(origin) |>
  slice_max(dep_delay, n = 1) |>
  relocate(dep_delay)
```

Which `tailnum` had the most flights?

```{r}
flights |>
  count(tailnum, sort = TRUE) |>
  slice(1:5)
```

### Grouping by multiple variables

You can create groups using more than one variable. For example, we could make a group for each date:

```{r}
daily <- flights |>
  group_by(year, month, day)
```

```{r}
#| message: true
daily |>
  summarise(number_of_rows = n())
```

When you summarize a tibble grouped by more than one variable, each summary peels off the last group, in this case `day`. It also shows a message. To suppress it, use `.groups = "drop_last"`:

```{r}
daily_flights <- daily |>
  summarise(
    n = n(),
    .groups = "drop_last"
  )
```

Alternatively, change the default behaviour by setting a different value, e.g. `"drop"` to drop all grouping or `"keep"` to preserve the same groups:

```{r}
daily |>
  summarise(n = n(), .groups = "keep") |>
  arrange(desc(n))
```

### Ungrouping

You can also remove grouping from a data frame without using `summarize()`:

```{r}
daily |>
  ungroup()
```

What happens when you summarize an ungrouped data frame?

```{r}
daily |>
  ungroup() |>
  summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    flights = n()
  )
```

You get a single row back, because dplyr treats all the rows in an ungrouped data frame as belonging to one group. It is basically the `flights` data as one group:

```{r}
flights |>
  summarise(
    av_delay = mean(dep_delay, na.rm = TRUE),
    flights = n()
  )
```

### .by

For instance:

```{r}
flights |>
  summarise(
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    total_flights = n(),
    .by = month
  )
```

or

```{r}
flights |>
  summarise(
    n = n(),
    .by = month
  )
```

This means: group the data by month, then summarise each group. With `.by`, you do both inside the one operation. `.by` is **per-operation grouping**.

You can also use `.by` with multiple variables:

```{r}
flights |>
  summarise(
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    total_flights = n(),
    .by = c(origin, dest)
  ) |>
  arrange(desc(avg_arr_delay))
```

For this summary, we group the flights by the combination of `origin` AND `dest`. Each unique origin + dest combination is a group.

Keep in mind:

- `group_by()` = group first, then do things.
- `.by` = group while doing this particular thing.

### Exercises

**1. Which carrier has the worst average delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about `flights |> group_by(carrier, dest) |> summarize(n())`.)**

```{r}
flights |>
  summarise(
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    avg_arr_delay = mean(arr_delay, na.rm = TRUE),
    .by = carrier,
    total_flight = n()
  ) |>
  arrange(desc(avg_dep_delay))
```

You don't have to compute multiple means.

```{r}
flights |>
  group_by(carrier, dest) |>
  summarize(n())
```

Part 2: the key is to look at carrier + destination together.

```{r}
flights |>
  summarise(
    avg_delay = mean(dep_delay, na.rm = TRUE),
    n = n(),
    .by = c(carrier, dest)
  ) |>
  arrange(dest)
```

**2. Find the flights that are most delayed upon departure from each destination.**

```{r}
flights |>
  group_by(dest) |>
  slice_max(dep_delay) |>
  arrange(dest, dep_delay) |>
  relocate(dep_delay, dest)
```

**3. How do delays vary over the course of the day? Illustrate your answer with a plot.**

Logic: group the flights by the hour they depart, calculate the average departure delay for each hour, and plot the averages to see how delays change throughout the day.

```{r}
flights |>
  summarise(
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    .by = hour
  ) |>
  ggplot(aes(hour, avg_dep_delay)) +
  geom_smooth()
```

A "go a little crazy" version:

```{r}
flights |>
  summarise(
    avg_dep_delay = mean(dep_delay, na.rm = TRUE),
    .by = hour
  ) |>
  ggplot(aes(hour, avg_dep_delay)) +
  geom_smooth(method = "lm", colour = "red") +
  labs(
    title = "Average Departure Delay By Hour",
    x = "Hour of Day",
    y = "Average Departure Delay (Minutes)"
  ) +
  theme_minimal()
```

```{r}
#| eval: false
ggsave(filename = "avg_delay plot.png")
```

**4. What happens if you supply a negative `n` to `slice_min()` and friends?**

```{r}
flights |>
  group_by(origin) |>
  slice_min(arr_delay, n = -1, with_ties = FALSE) |>
  count()
```

For each origin, it finds the flight with the minimum `arr_delay`, removes it, and keeps everything else. That is what a negative `n` does: it means "remove that many of the smallest/largest rows". `slice_head()` and `slice_tail()` also accept a negative `n`, but there the meaning is based on row position rather than on minimum/maximum values.

**5. Explain what `count()` does in terms of the dplyr verbs you just learned. What does the `sort` argument to `count()` do?**

`count()` is a shortcut for `group_by()` + `summarise(n = n())`. `sort = TRUE` sorts the resulting counts from largest to smallest.

```{r}
flights |>
  count(origin)
```

This is basically:

```{r}
flights |>
  group_by(origin) |>
  summarise(n = n())
```

With `sort = TRUE`:

```{r}
flights |>
  count(origin, sort = TRUE)
```

**6. Suppose we have the following tiny data frame:**

```{r}
df <- tibble(
  x = 1:5,
  y = c("a", "b", "a", "a", "b"),
  z = c("K", "K", "L", "L", "K")
)
```

**a. Write down what you think the output will look like, then check if you were correct, and describe what `group_by()` does.**

```{r}
df |>
  group_by(y)
```

`group_by()` does NOT collapse or rearrange the data. R now remembers "this data is grouped according to `y`". So there are two groups: `y = "a"` (rows 1, 3, 4) and `y = "b"` (rows 2, 5). It says: "From now on, when I use a verb that works with groups, treat the data separately according to `y`."

```{r}
df |>
  group_by(y) |>
  summarise(n = n())
```

`group_by(y)` doesn't change the rows or columns. It adds grouping information so that subsequent dplyr verbs can operate separately for each value of `y`.

**b. Write down what you think the output will look like, then check if you were correct, and describe what `arrange()` does. Also comment on how it's different from the `group_by()` in part (a).**

```{r}
df |>
  arrange(y)
```

`arrange(y)` sorts the rows of the tibble according to the values of `y`, putting them in alphabetical order. The other variables move with their corresponding rows, so the values in each row stay together.

Unlike `group_by()`, `arrange()` actually changes the order of the rows. `group_by()` does not rearrange the data; it only adds grouping information for subsequent operations.

**c. Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does.**

```{r}
df |>
  group_by(y) |>
  summarise(mean_x = mean(x))
```

The pipeline first groups the data by `y`, creating groups for a and b. `summarise()` then calculates the mean of `x` separately for each group, producing one row for each value of `y`. The pipe `|>` passes the result of each step to the next function.

**d. Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does. Then comment on what the message says.**

```{r}
#| message: true
df |>
  group_by(y, z) |>
  summarise(mean_x = mean(x))
```

`group_by(y, z)` tells `summarise()` to calculate the summaries for each y-z combination. After summarising, `summarise()` drops the last grouping level by default, leaving the result grouped by `y`. That is what the message says.

**e. Write down what you think the output will look like, then check if you were correct, and describe what the pipeline does. How is the output different from the one in part (d)?**

```{r}
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x), .groups = "drop")
```

`.groups = "drop"` removes all grouping after `summarise()`. The actual summary results are the same as in part (d), but the resulting tibble is completely ungrouped. `.groups` doesn't change how the mean is calculated here; it controls what happens to the grouping after the calculation.

**f. Write down what you think the outputs will look like, then check if you were correct, and describe what each pipeline does. How are the outputs of the two pipelines different?**

```{r}
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x))
```

```{r}
df |>
  group_by(y, z) |>
  mutate(mean_x = mean(x))
```

`summarise()` reduces each group to one row, while `mutate()` keeps every original row and adds the group-level calculation to each row.

## Case study: aggregates and sample size

If you haven't installed the Lahman package, run this once in the Console:

```{r}
#| eval: false
install.packages("Lahman")
```

```{r}
library(Lahman)
```

```{r}
#| eval: false
View(LahmanData)
View(Teams)
View(Batting)
```

When you summarise data, always include `n()` so you know how many observations your conclusion is based on. A conclusion based on very little data is unreliable and potentially misleading. The baseball example shows this: a player's hit rate means very little if they only had a few attempts.

Here we compare the proportion of times a player gets a hit (`H`) with the number of times they try to put the ball in play (`AB`):

```{r}
batters <- Lahman::Batting |>
  group_by(playerID) |>
  summarize(
    performance = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
    n = sum(AB, na.rm = TRUE)
  )
```

The `::` basically means "take `Batting` from the Lahman package".

When plotting performance against sample size, players with fewer observations show greater variation, while performance becomes more stable as the number of observations increases. There is also a positive correlation between performance and opportunities, because teams tend to give better-performing players more opportunities.

```{r}
batters |>
  filter(n > 100) |>
  ggplot(aes(x = n, y = performance)) +
  geom_point(alpha = 1 / 10) +
  geom_smooth(se = FALSE)
```

Ranking players only by batting average can be misleading, because very high averages may be based on very few at-bats. Sample size should therefore be considered when interpreting or ranking performance.

```{r}
batters |>
  arrange(desc(performance))
```
