---
execute:
  warning: false
  message: false
---

# Data tidying {#sec-data-tidying}

## Introduction

### Prerequisites

```{r}
library(tidyverse)
```

## Tidy data

You can represent the same underlying data in multiple ways. Tidy data is easier to read and work with.

There are three interrelated rules that make a dataset tidy:

1. Each variable is a column; each column is a variable.
2. Each observation is a row; each row is an observation.
3. Each value is a cell; each cell is a single value.

The tidyverse packages are designed to work with tidy data. Example:

```{r}
table1
```

Compute the rate per 10,000:

```{r}
table1 |>
  mutate(rate = cases / population * 10000)
```

Total cases per year:

```{r}
table1 |>
  summarise(
    total_cases = sum(cases),
    .by = year
  )
```

Visualize changes over time:

```{r}
ggplot(table1, aes(x = year, y = cases)) +
  geom_line(aes(group = country), color = "grey50") +
  geom_point(aes(color = country, shape = country)) +
  scale_x_continuous(breaks = c(1999, 2000)) # x-axis breaks at 1999 and 2000
```

### Exercises

**1. For each of the sample tables, describe what each observation and each column represents.**

```{r}
table1
table2
table3
```

- `table1`: each row represents a country in a particular year. Cases and population are stored in separate columns.
- `table2`: each row represents a country, year, and measurement type. Cases and population are stored in one `count` column, which makes this a longer format.
- `table3`: each row represents a country in a particular year. Cases and population are combined into a character column called `rate`, which makes numerical calculations more difficult.

**2. Sketch out the process you'd use to calculate the rate for `table2` and `table3`. You will need to perform four operations:**

a. Extract the number of TB cases per country per year.
b. Extract the matching population per country per year.
c. Divide cases by population, and multiply by 10000.
d. Store back in the appropriate place.

From ChatGPT:

> For `table2`, I would group the data by country and year, extract the count for cases and the count for population, divide cases by population, multiply by 10,000, and store the result in a new `rate` column. For `table3`, I would first separate the `rate` column into cases and population, convert both values into numbers, calculate cases divided by population times 10,000, and replace the original text with the numerical rate.

The code for `table2`:

```{r}
table2 |>
  pivot_wider(
    names_from = type,
    values_from = count
  ) |>
  mutate(
    rate = cases / population * 10000
  )
```

## Lengthening data

Pivoting transforms data into a tidy form: variables in columns and observations in rows.

The tidyr functions for this are:

1. `pivot_longer()`
2. `pivot_wider()`

### Data in column names

```{r}
billboard
```

In the Billboard data, each row initially represents one song. The weekly rankings are spread across the columns `wk1` to `wk76`. The column names contain the week variable, while the cell values contain the rank variable. `pivot_longer()` converts these weekly columns into rows, so that each row represents one song's rank in one week.

```{r}
billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "position"
  )
```

Or a different way of selecting the columns:

```{r}
billboard |>
  pivot_longer(
    cols = !c(artist, track, date.entered),
    names_to = "week",
    values_to = "rank"
  )
```

`pivot_longer()` converts the columns beginning with `wk` into rows. The old column names are stored in a new `week` column, while the ranking values are stored in a new `rank` column.

- `cols` identifies the columns to pivot.
- `names_to` gives a name to the new column containing the old column names.
- `values_to` gives a name to the new column containing the old cell values.

**Structural missing values** occur because of how the data is organized, rather than because an expected value is unknown. In the Billboard data, `NA` means a song was not in the Top 100 during that week. `values_drop_na = TRUE` removes these unnecessary `NA` rows when using `pivot_longer()`.

```{r}
billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    values_to = "position",
    names_to = "week",
    values_drop_na = TRUE
  )
```

To make computation easier, we can convert the values of `week` from character strings to numbers, using `mutate()` and `readr::parse_number()`:

```{r}
billboard_longer <- billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "position",
    values_drop_na = TRUE
  ) |>
  mutate(week = parse_number(week))

billboard_longer
```

`pivot_longer()` initially creates `week` as character values such as "wk1" and "wk2". `mutate()` with `parse_number()` extracts the numbers and converts `week` to a numeric variable, making future calculations easier.

We can now visualize how song ranks vary over time:

```{r}
billboard_longer |>
  ggplot(aes(week, position, group = track)) +
  geom_line(alpha = 0.25) +
  scale_y_reverse()
```

### How does pivoting work?

Suppose we have 3 patients, each with 2 blood pressure readings:

```{r}
df <- tribble(
  ~id,  ~bp1, ~bp2,
  "A",  100,  120,
  "B",  140,  115,
  "C",  120,  125
)
```

We want our new dataset to have three variables: `id` (which already exists), `measurement` (the column names), and `value` (the cell values). To achieve this, we pivot `df` longer:

```{r}
df |>
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
```
