---
execute:
  warning: false
  message: false
---

# Workflow: basics {#sec-workflow-basics}

## Coding basics

Using R to do basic math calculations:

```{r}
1 / 200 * 30
(59 + 323 + 2) / 3
sin(pi / 3)
```

Create new objects with the assignment operator `<-` (shortcut: Alt + -):

```{r}
x <- 3 * 5
```

Note that the value of `x` is not printed, it's just stored. To view the value, type `x` in the console.

```{r}
x
```

You can combine multiple elements into a vector with `c()`:

```{r}
primes <- c(2, 3, 5, 7, 11, 13)
```

Basic arithmetic on vectors is applied to every element of the vector:

```{r}
primes * 2
primes - 1
```

All R statements where you create objects (assignment statements) have the same form:

```{r}
object_name <- 58
```

## Comments

Use comments to explain the **why** of your code, not the how or the what.

## What's in a name?

Object names must start with a letter and can only contain letters, numbers, `_` and `.`.

```{r}
#| eval: false
i_use_snake_case
otherPeopleUseCamelCase
some.people.use.periods
And_aFew.People_RENOUNCEconvention
```

`snake_case` is better: lowercase only, with `_` between words.

To inspect an object, type its name:

```{r}
x
this_is_a_long_ass_name <- 2.6
```

To inspect a long name, type the first few letters (like "this") and press Tab to autocomplete.

Another assignment:

```{r}
r_rocks <- 3 * 3
```

## Calling functions

R has built-in functions that are called like this:

```r
function_name(argument1 = value1, argument2 = value2, ...)
```

```{r}
seq(from = 1, to = 10)
```

The names of the first several arguments are often omitted, so we can rewrite this as:

```{r}
seq(1, 10)
```

```{r}
x <- "hello world"
x
```

## Exercises

**1. Why does this code not work?**

```{r}
#| error: true
my_variable <- 10
my_varıable
```

The second line has a typo: the "i" in `variable` is a different character (a dotless ı), so R looks for an object that doesn't exist.

**2. Tweak each of the following R commands so that they run correctly.**

The broken versions:

```{r}
#| eval: false
libary(todyverse)

ggplot(dTA = mpg) +
  geom_point(maping = aes(x = displ y = hwy)) +
  geom_smooth(method = "lm)
```

The tweaked versions:

```{r}
library(tidyverse)

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(method = "lm")
```

But I can write it more concisely (at my level, lol):

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth(method = "lm")
```

**3. Press Option + Shift + K / Alt + Shift + K. What happens? How can you get to the same place using the menus?**

The keyboard shortcut quick reference opens. Using the menus: Help → Keyboard Shortcuts Help.

**4. Revisit an exercise from Section 1.6. Run the following lines of code. Which of the two plots is saved as `mpg-plot.png`? Why?**

```{r}
my_bar_plot <- ggplot(mpg, aes(x = class)) +
  geom_bar()
my_scatter_plot <- ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
```

```{r}
#| eval: false
ggsave(filename = "mpg-plot.png", plot = my_bar_plot)
```

The bar plot is saved, because `plot = my_bar_plot` tells `ggsave()` exactly which plot to use.

Note that when a plot is assigned to an object, it is stored but **not shown**. Call the object by name to see it:

```{r}
my_scatter_plot
```

```{r}
my_bar_plot
```
